Statistical Relational Models as Polyadic Measure Algebras

### 2020-03-02 paper.tex
Proof that a measure can be transferred from the BA to the PA

\begin{definition}
  A \emph{Boolean algebra} is a tuple $(\mathbf{B}, \land, \lor, \neg, 0, 1)$ of
  a set $\mathbf{B}$ with operations $\land, \lor, \neg$ and elements $0, 1 \in
  \mathbf{B}$ such that the following axioms hold for all $a, b, \in
  \mathbf{B}$:
  \begin{itemize}
  \item both $\land$ and $\lor$ are associative and commutative;
  \item $a \lor (a \land b) = a$, and $a \land (a \lor b) = a$;
  \item $0$ is the identity of $\lor$, and $1$ is the identity of $\land$;
  \item $\lor$ distributes over $\land$ and vice versa;
  \item $a \lor \neg a = 1$, and $a \land \neg a = 0$.
  \end{itemize}

@@

\begin{definition}[\cite{DBLP:books/daglib/0090259}]
  An element $a \ne 0$ of a Boolean algebra $\mathbf{B}$ is an \emph{atom} if
  there is no $x \in \mathbf{B}$ such that $0 < x < a$. A Boolean algebra is
  \emph{atomic} if for every $a \in \mathbf{B} \setminus \{ 0 \}$, there is an
  atom $x$ such that $x \le a$.
\end{definition}

@@

% Defining E
For any $J \subseteq I$, let $J_*$ be the relation on $X_I$ defined by
\[
  xJ_*y \quad \iff \quad x_i = y_i \quad \text{for all } i \in I \setminus J
\]
for all $x, y \in X^I$. For any $J \subseteq I$, we then define $\bm\exists(J)$
to be a transformation $\mathbf{A^*} \to \mathbf{A^*}$ defined by
\[
  \bm\exists(J)p(x) = \bigvee_{\substack{y \in X^I,\\ xJ_*y}} p(y)
\]
for all $p \in \mathbf{A^*}$, provided this supremum exists for all $x \in
X^I$\footnote{The universal quantifier $\bm\forall(J)$ is then defined as
  $\bm\forall(J)p = \neg(\bm\exists(J)\neg p)$ for all $p \in \mathbf{A^*}$.}.

@@

\begin{proposition}
  Let $\mathbf{B}$ be a finite Boolean algebra with a measure $m :
  \mathbf{B} \to [0, 1]$. Let $\mathbf{A}$ be a $\mathbf{B}$-valued functional
  polyadic algebra with domain $X$ and variables $I$. For any $p \in
  \mathbf{A}$, let $\sup(p) = \bigvee_{x \in X^I} p(x)$. Let $m^* : \mathbf{A}
  \to [0, 1]$ be defined by
  \[
    m^*(p) = \sum_{\substack{\text{atom } y \in \mathbf{B}:\\ y \le \sup(p)}}
    m(y)
  \]
  for all $p \in \mathbf{A}$. Then $m^*$ is a measure on $\mathbf{A}$.
\end{proposition}

\begin{remark}
  While defining $m^*$ as $m^*(p) = m(\sup(p))$ might look tempting, this
  definition is not additive.
\end{remark}

\begin{proof} % TODO: update the proof
  First, since $1 \in \mathbf{A}$ is a function $1 : X^I \to \mathbf{B}$ defined
  as $1(x) = 1$, we have that $m^*(1) = m(1) = 1$. If $m^*(p) = 0$ for some $p
  \in \mathbf{A}$, then it must be the case that
  \[
    \bigvee_{x \in X^I} p(x) = 0
  \]
  which means that for all $x \in X^I$, $p(x) = 0$, which is equivalent to
  saying that $p = 0$. Thus, $m^*(p) > 0$ for $p \in \mathbf{A} \setminus \{ 0
  \}$.
\end{proof}

% TODO: how to interpret first-order sentences in this algebra
% TODO: define WMC: connect my definition of WMC with the original one
% TODO: condition for a measurable BA to be representable as WMC
% TODO: can I show that WMC is the only possible probability measure?
% TODO: abstraction refinements as homomorphisms.
% TODO: definition of a measure-preserving homomorphism from Thomas Jech
% TODO: a Boolean algebra is approximable if its Stone space is approximable

>>>>

\begin{definition} \label{def:ba}
  A \emph{Boolean algebra} is a tuple $(\mathbf{B}, \land, \lor, \neg, 0, 1)$ of
  a set $\mathbf{B}$ with operations $\land, \lor, \neg$ and elements $0, 1 \in
  \mathbf{B}$ such that the following axioms hold for all $a, b, \in
  \mathbf{B}$:
  \begin{itemize}
  \item both $\land$ and $\lor$ are associative and commutative;
  \item $a \lor (a \land b) = a$, and $a \land (a \lor b) = a$;
  \item $0$ is the identity of $\lor$, and $1$ is the identity of $\land$;
  \item $\lor$ distributes over $\land$ and vice versa;
  \item $a \lor \neg a = 1$, and $a \land \neg a = 0$.
  \end{itemize}

@@

% TODO: Which definition do I actually need?
\begin{definition}[\cite{DBLP:books/daglib/0090259,levasseur2012applied}]
  An element $a \ne 0$ of a Boolean algebra $\mathbf{B}$ is an \emph{atom} if
  there is no $x \in \mathbf{B}$ such that $0 < x < a$. Equivalently, $a \ne 0$
  is an atom if, for all $x \in \mathbf{B}$, either $x \land a = a$ or $x \land
  a = 0$. A Boolean algebra is \emph{atomic} if for every $a \in \mathbf{B}
  \setminus \{0 \}$, there is an atom $x$ such that $x \le a$.
\end{definition}

@@

% Defining E
For any $J \subseteq I$, let $J_*$ be the relation on $X^I$ defined by
\[
  xJ_*y \quad \iff \quad x_i = y_i \quad \text{for all } i \in I \setminus J
\]
for all $x, y \in X^I$. For any $J \subseteq I$, we then define $\bm\exists(J)$
to be a transformation $\mathbf{A^*} \to \mathbf{A^*}$ defined by
\[
  \bm\exists(J)p(x) = \bigvee_{\substack{y \in X^I,\\ xJ_*y}} p(y)
\]
for all $p \in \mathbf{A^*}$, provided this supremum exists for all $x \in
X^I$\footnote{The universal quantifier $\bm\forall(J)$ is then defined as
  $\bm\forall(J)p = \neg(\bm\exists(J)\neg p)$ for all $p \in \mathbf{A^*}$.}.

@@

\begin{proposition}
  Let $\mathbf{B}$ be a finite Boolean algebra with a measure $m :
  \mathbf{B} \to [0, 1]$. Let $\mathbf{A}$ be a $\mathbf{B}$-valued functional
  polyadic algebra with domain $X$ and variables $I$. Let $m^* : \mathbf{A} \to
  \mathbb{R}_{\ge 0}$ be defined by
  \[
    m^*(p) = \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le p(x)}} m(y)
  \]
  for all $p \in \mathbf{A}$. Then $m^*$ is a measure on $\mathbf{A}$.
\end{proposition}

\begin{remark}
  While definitions of $m^*$ such as
  \[
    m^*(p) = m \left( \bigvee_{x \in X^I} p(x) \right)
  \]
  might look tempting, they are not additive.
\end{remark}

\begin{proof} % TODO: update the proof
  First, we can show that $m^*(1) = 1$ by observing that
  \[
    m^*(1) = \sum_{\text{atoms } y \in \mathbf{B}} m(y) = m \left(
      \bigvee_{\text{atoms } y \in \mathbf{B}} y \right) = m(1) = 1,
  \]
  where we use \cref{thm:representation} and express $1 \in \mathbf{B}$ as the
  supremum of all atoms in $\mathbf{B}$ \cite{ganesh2006introduction}. Clearly
  $m^*(p) \ge 0$ for all $p \in \mathbf{A}$, so we can restrict the codomain of
  $m^*$ to $[0, 1]$.

  Next, we want to show that $m^*(p) > 0$ for all $p \in \mathbf{A} \setminus \{
  0 \}$. If $p \ne 0$, then there must be some $x' \in X^I$ such that $p(x') \ne
  0$. But then, since finite Boolean algebras are atomic, there must also be an
  atom $y \in \mathbf{B}$ such that $y \le p(x')$. Therefore, $m^*(p) \ge m(y) >
  0$, finishing this part of the proof.

  Let $p, q \in \mathbf{A}$ be such that $p \land q = 0$. We want to show
  that $m^*(p \lor q) = m^*(p) \lor m^*(q)$. First, note that
  \[
    y \le (p \lor q)(x) = p(x) \lor q(x)
  \]
  if and only if
  \[
    y = (p(x) \lor q(x)) \land y = (p(x) \land y) \lor (q(x) \land y)
  \]
  by \cref{def:ba}. Also note that
  \[
    (p(x) \land y) \land (q(x) \land y) = p(x) \land q(x) \land y = (p \land
    q)(x) \land y = 0 \land y = 0,
  \]
  so
  \[
    m(y) = m((p(x) \land y) \lor (q(x) \land y)) = m(p(x) \land y) + m(q(x) \land y)
  \]
  by \cref{def:measure} which then leads to
  \begin{align*}
    m^*(p \lor q) &= \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x)}} m(y) = \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x)}} m(p(x) \land y) + m(q(x) \land y) \\
                  &= \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x)}} m(p(x) \land y) + \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x)}} m(q(x) \land y).
  \end{align*}
  Since $y$ is an atom,
  \[
    p(x) \land y  = \begin{cases}
      y & \text{if } y \le p(x) \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  so
  \begin{align*}
    m^*(p \lor q) &= \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x) \text{ and } y \le p(x)}} m(p(x) \land y) + \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{s.t.}\\ \exists x \in X^I:\, y \le (p \lor q)(x) \text{ and } y \le q(x)}} m(q(x) \land y) \\
                  &= \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le p(x)}} m(p(x) \land y) + \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{s.t.}\\ \exists x \in X^I:\, y \le q(x)}} m(q(x) \land y) \\
                  &= \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le p(x)}} m(y) + \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{s.t.}\\ \exists x \in X^I:\, y \le q(x)}} m(y) = m^*(p) + m^*(q),
  \end{align*}
  finishing the proof that $m^*$ is a measure.
\end{proof}

% TODO: how to interpret first-order sentences in this algebra
% TODO: define WMC: connect my definition of WMC with the original one
% TODO: condition for a measurable BA to be representable as WMC
% TODO: can I show that a generalisation of WMC is the only possible probability measure (when it exists)?
% TODO: abstraction refinements as homomorphisms.
% TODO: definition of a measure-preserving homomorphism from Thomas Jech
% TODO: a Boolean algebra is approximable if its Stone space is approximable
